---
title: "Lab 3: Data Wrangling for Visualization"
author: Kristen McDaniel
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
    code-tools: true
execute:
  eval: false
  warning: false
  message: false
---

# Setup

```{r, load-packages}

library(tidyverse)

```

# Part 1: Create Raw Data

```{r, raw-data}

raw_health_data <- data.frame(
  county = c(
    "Fulton", "DeKalb", "Cobb", "Gwinnett",
    "Clayton", "Henry", "Cherokee", "Fulton", "Douglas"
  ),
  region = c(
    "Central", "Central", "North", "North",
    "South", "South", "North", "Central", "West"
  ),
  population_2023 = c(
    1063937, 764597, 776781, 975353,
    303435, 245684, 316837, 1063937, 147843
  ),
  cases_2023 = c(
    8420, 5210, 4385, 6120,
    2745, 1987, 1765, 8420, 1248
  ),
  population_2024 = c(
    1072000, 770100, 783500, 987600,
    307200, 251300, 323500, 1072000, 151200
  ),
  cases_2024 = c(
    8650, 5385, 4520, 6355,
    2810, 2075, NA, 8650, 1315
  )
)

raw_health_data

```

# Part 2: Diagnose

```{r, diagnose-data}

glimpse(raw_health_data)

```

# Part 3: Remove Duplicates

```{r, remove-duplicates}

health_data_clean <- raw_health_data %>%
  distinct(county, .keep_all = TRUE)

```

# Part 4: Reshape Long

```{r, reshape-long}

health_data_clean1 <- health_data_clean %>% 
  pivot_longer(cols = c(population_2023, cases_2023, population_2024, cases_2024),
    names_to = c(".value", "year"),
    names_pattern = "(population|cases)_(2023|2024)"
  )

```

# Part 5: Calculate Rates

```{r, calculate-rates}

health_data_clean2 <- health_data_clean1 %>%
  mutate(
    cases_thousands = cases / 1e3
  )

```

# Part 6: Summarize

```{r, summarize-region}

mean_rate <- health_data_clean2 %>% group_by(region, year) %>%
  summarise(
    mean_cases = mean(cases_thousands, na.rm = TRUE)
  )

```

# Part 7: Join Contextual Data

```{r}

rural_status <- data.frame(
  county = c("Fulton", "DeKalb", "Cobb", "Gwinnett",
             "Clayton", "Henry", "Cherokee", "Douglas"),
  rural_status = c("Urban", "Urban", "Urban", "Suburban",
                   "Urban", "Suburban", "Suburban", "Rural")
)

# check duplicates in rural-status table
rural_status %>%
  count(county) %>%
  filter(n > 1)

# join contextual data
health_data_joined <- health_data_clean2 %>%
  left_join(rural_status, by = "county")

# check unmatched rows
health_data_joined %>%
  filter(is.na(rural_status)) %>%
  select(county)

# check row counts before and after join
nrow(health_data_clean2)
nrow(health_data_joined)

# joined data
health_data_joined

```

# Part 8: Visualize


```{r, line-graph}

line_graph <- ggplot(
  mean_rate,
  aes(
    x = year,
    y = mean_cases,
    color = region,
    group = region
   ) 
  )+
    geom_line(
  ) +
  geom_point(
    size = 2,
    alpha = 0.5
  ) +
  labs(
    title = "Mean Health Rate by Region and Year ",
    x = "Year",
    y = "Mean Cases",
    caption = "Data Source: Simulated"
    ) +
    theme_minimal()

line_graph

```

```{r, line-save}

ggsave(
  filename = "../Figures/lab03_line_graph.png",
  plot = line_graph,
  width = 8,
  height = 6,
  dpi = 300
)

```


# Ranked 2024 county graph

```{r, 2024-graph}

mean_rate1 <- health_data_clean2 %>% group_by(county, year) %>%
  summarise(
    mean_cases = mean(cases_thousands, na.rm = TRUE)
  )

hd2024 <- mean_rate1 %>%
  filter(year == 2024) %>%
  ggplot(
    aes(
      x = reorder(county, mean_cases),
      y = mean_cases,
      color = county
    )
  ) +
  geom_col() +
  coord_flip() +
  labs(
    title = "2024 Mean Cases by County",
    x = "County",
    y = "Mean Cases"
  ) +
  theme_minimal()

hd2024

```
```{r, 2024-save}

ggsave(
  filename = "../Figures/lab03_hd2024.png",
  plot = hd2024,
  width = 8,
  height = 6,
  dpi = 300
)

```


# Rural-status comparison

```{r, rural-status}

mean_rate2 <- health_data_joined %>% group_by(county,rural_status, year) %>%
  summarise(
    mean_cases = mean(cases_thousands, na.rm = TRUE)
  )

rural_status_graph <- ggplot(
  mean_rate2, 
       aes(
         x = rural_status, 
         y = mean_cases,
         color = rural_status
         )
  ) +
  geom_boxplot() +
  labs(
    title = "Health Rate by Rural Status",
    x = "Rural Status",
    y = "Mean Cases"
  ) +
  theme_minimal()

rural_status_graph

```
```{r, boxplot-save}

ggsave(
  filename = "../Figures/lab03_rural_status_graph.png",
  plot = rural_status_graph,
  width = 8,
  height = 6,
  dpi = 300
)

```



# Reflection

Explain why long format was useful, what causes row multiplication during joins, and why rates are preferable to counts.

**Long format was useful, because it made the data easier to analyze and visualize across different years. Row multiplication during joins can occur when there are duplicate county names or multiple matching rows in the table being joined. Rates are preferable counts, because they account for differences in population size, which allows for more meaningful comparisons.**
