---
title: "Lab 1: Reproducible Data Visualization Workflow"
format:
  html:
    toc: true
    code-copy: true
    code-tools: true
execute:
  warning: false
  message: false
---

## Load Packages

```{r load-packages}

#install.packages('tidyverse')
  library(tidyverse)

#install.packages('here')
  library(here)

#install.packages('fs')
  library(fs)

#install.packages('readr')
  library(readr)

#install.packages('dplyr')
  library(dplyr)

```

## Project Root

```{r project-root}

here()

```

Answer:

1. What path was printed?
  "/Users/kmcdaniel/Desktop/DataVisualization"
  
2. Does it point to the correct course project?
  Yes
  
3. Why is working from an R Project preferable to manually setting the working directory?
  Easier and more efficient


## Inspect the Folder Structure

```{r inspect-structure}

dir_ls()

```

## Create a Simulated Public Health Dataset

Create a reproducible county-level dataset with the following variables:

- `county`
- `region`
- `adult_population`
- `asthma_visits`
- `vaccination_rate`
- `asthma_visit_rate`

Requirements:

- Include 12 fictional counties.
- Include three regions: North, Central, and South.
- Use `set.seed(716)`.
- Generate plausible values for population, visits, and vaccination coverage.
- Calculate asthma visit rate per 100,000.
- Round the rate to one decimal place.
- Use tidy formatting.

```{r simulation}

set.seed(716)

# Categorical Data

county_health_data <- tibble(

county = c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L"),
region = c("North", "South", "Central", "Central", "North", "North", "South", "Central", "South", "South", "North", "Central"),
adult_population = c("42,315", "67,842", "31,576", "89,421", "54,738", "112,605", "76,193", "38,947", "95,284", "61,729", "47,856", "83,514" ),
asthma_visits = c( "1,284", "2,176", "947", "3,052", "1,693", "3,841", "2,487", "1,126", "3,317", "1,954", "1,438", "2,731"),
vaccination_rate = c("68.4", "72.1", "75.3", "69.8", "78.6", "73.5", "64.2", "76.8", "70.6", "66.9", "74.1", "85.3"),
asthma_visit_rate = c("30.35", "32.07", "29.99", "34.13", "30.93", "34.11", "32.64", "28.91", "34.81", "31.65", "30.05", "32.70")
) %>%
  mutate(
    adult_population = as.numeric(gsub(",", "", adult_population)),
    asthma_visits    = as.numeric(gsub(",", "", asthma_visits)),
    vaccination_rate = as.numeric(vaccination_rate),
    asthma_visit_rate = as.numeric(asthma_visit_rate)
  )

```


## Save Dataset

```{r dataset}

write_csv(county_health_data, here("Data", "county_health_data.csv"))
  
```


# Part 9: Remove and Re-Import the Dataset


```{r remove/import-dataset}

rm(county_health_data)

county_health_data <- read_csv(here("Data", "county_health_data.csv"))

```

## Inspect the Data

```{r inspect-dataset}

glimpse(county_health_data)

```
## One Row Summary

```{r onerow-summary}

county_number <- nrow(county_health_data)
  # number of counties = 12

vr_mean <- mean(county_health_data$vaccination_rate)
  # vaccination rate mean = 72.96667

avr_median <- median(county_health_data$asthma_visit_rate)
  # asthma visit mean = 31.86

one_row <- data.frame(county_number, vr_mean, avr_median)
  print(one_row)

```

Answer:

1. What does one row represent?
  One observation
  
2. How many variables are included?
  3
  
3. Which variables are categorical?
  County

4. Which variables are numeric?
  Vaccination rates and Asthma visit rates
  
5. Which variable should be used to compare asthma burden fairly across counties?
  Asthma visit rate

## Ranked Vaccination Bar Chart

Create a horizontal bar chart ranking counties by vaccination coverage.

Requirements:

- Reorder counties by vaccination rate.
- Add a dashed reference line at 80%.
- Use an informative title and subtitle.
- Label the y-axis as vaccination coverage.
- State that the values are simulated.
- Use a readable theme.
- Store the graph as `vaccination_plot`.

```{r vaccination-plot}

vaccination_plot <- ggplot(
  county_health_data,
  aes(
    x = reorder(
      county, vaccination_rate
                ),
    y = vaccination_rate
    )
  ) +
  geom_col(
    fill= "blue"
    ) +
  geom_hline(
    yintercept = 80, 
             linetype = "dashed"
    ) +
  labs(
    title = "Vaccination Coverage Across Counties",
    subtitle = "County vaccination rates compared with the 80% reference level",
    x = "County",
    y = "Vaccination Coverage",
    caption = "Data Source: Simulated"
    ) +
  theme_minimal()

vaccination_plot

```

## Interpret the Figure

1. What pattern does the figure show?
2. Which counties appear furthest below the target?
3. What context would be required before using this graph for a real decision?

The figure shows that vaccination coverage varies across counties, with rates ranging from about 64% to 85%. Most counties are below the 80% target, while only County L exceeds it. Context, such as available resources, access to care, socioeconomic makeup, type of area, and other community factors, should be considered.

## Save the Vaccination Figure


```{r vaccination-figure}

ggsave(
  filename = "../Figures/lab01_vaccination_coverage.png",
  vaccination_plot,
  width = 8,
  height = 6,
  dpi = 300
)

```

## Independent Visualization

- Scatterplot of adult population and asthma visits,

```{r independent-plot}

independent_plot <- ggplot(
  county_health_data,
  aes(
    x = adult_population,
    y = asthma_visits
    )
  ) +
  geom_point(
    color = "#2b5c8f", 
             size = 4, 
             alpha = 0.8
    ) +
  geom_smooth(
    method = "lm", 
              se = FALSE
    ) +
  geom_text(
    aes(
      label = county), 
       vjust = -1, 
       size = 3
    ) +
  labs(
    title = "Adult Population and Asthma Visits by County",
    subtitle = "County-level comparison of population size and asthma visits",
    x = "Adult Population",
    y = "Asthma Visits",
    caption = "Data Source: Simulated"
    ) +
  theme_minimal(
    base_size = 10
    ) +
  theme(
    plot.title = element_text(
      face = "bold", 
      size = 10
      ),
    plot.subtitle = element_text(
      color = "grey40", 
      margin = margin(b = 5),
      size = 9
      ),
    panel.grid.minor = element_blank()
    
  )

independent_plot

```

## Save the Independent Figure

```{r independent-figure}

ggsave(
  filename = "../Figures/lab01_independent_visualization.png",
  independent_plot,
  width = 8,
  height = 5,
  dpi = 300
)

```

# Reflection Questions

Answer each question in complete sentences.

## Question 1

Why is reproducibility important in public health research and practice?

**Response:**  

## Question 2

How does a consistent project structure improve collaboration and review?

**Response:**  

## Question 3

Why are relative paths preferable to absolute paths?

**Response:**  

## Question 4

How does GitHub contribute to reproducibility?

**Response:**  

## Question 5

What was the most challenging part of the lab?

**Response:**  

# Deliverables

Submit or push:

- [ ] completed student lab QMD
- [ ] rendered HTML file
- [ ] `data/county_health_data.csv`
- [ ] `figures/lab01_vaccination_coverage.png`
- [ ] `figures/lab01_independent_visualization.png`
- [ ] completed reflections
- [ ] meaningful Git commit
- [ ] updated GitHub repository

# Submission Checklist

Before submitting, confirm:

- [ ] the project opens from the `.Rproj` file
- [ ] the QMD renders without errors
- [ ] all required folders exist
- [ ] the dataset is inside `data/`
- [ ] relative paths are used
- [ ] both visualizations display correctly
- [ ] both figures are saved inside `figures/`
- [ ] reflections are complete
- [ ] work is committed and pushed
- [ ] no private or sensitive information was uploaded

# Troubleshooting

# Dataset Cannot Be Found

Check:

- file location,
- spelling,
- capitalization,
- project root,
- and use of `here()`.

# Package Is Missing

Install the package in the Console:

```r
install.packages("package_name")
```

Do not place package-installation commands in the final lab.

# Quarto Will Not Render

Read the first error message and check:

- commas,
- parentheses,
- quotation marks,
- object names,
- and package availability.

# Git Pane Is Missing

Confirm:

- Git is installed,
- the project is a Git repository,
- and the correct project is open.

# Stretch Challenge

Improve one graph by adding one of the following:

- direct labels,
- a colorblind-friendly palette,
- an annotation,
- or a meaningful highlight.

Write two sentences explaining why the change improves communication.

# Looking Ahead

In Lab 2, you will use the Grammar of Graphics to build and compare several visualization types for different public health questions.
